Investigating shape standardization using geometric morphometry of pottery from Iron Age northeastern Taiwan

### 2020-04-11 analysis/paper/paper.Rmd
combine tps and PCA plots into one figure

```{r jpg-coordinates, eval=FALSE}
# start shape exploration
library(Momocs)
#  it freezes on the empty ones...
jpgs_imported <- import_jpg(jpgs, threshold = 0.5)

@@

We found that 13 harmonics captured 99% of the total harmonic power in our elliptic Fourier coefficients of `r total_no` vessels from three phases. Figure \@ref(fig:thin-plate-splines) shows vessel shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the most greatest differences between pre-European and Chinese periods.

```{r thin-plate-splines, fig.cap= "Comparison of average vessel shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another."}
# Thin plate spline analysis (TPS)
# visualize the outline deformations required to pass from an extreme of the morphospace to the other
# Deformation grids

@@

par(mfrow=c(1,2)) # plot them side by side

tps_grid(`Pre-European`, `Post-European`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
           amp = 1, grid.size = 18, legend = T)

tps_grid(`Pre-European`, `Chinese`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
           amp = 1, grid.size = 18, legend = T)

# Isodeformation lines
png(here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)

@@

The first two principal components (PCs) of the PCA on the elliptic Fourier coefficients explain `r vari_two`% of the total variance, of which `r pc1`% is explained by the first principle component. With the third component, the first three principal components explain `r vari_three`% of the total variance. PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped (Figure \@ref(fig:first-two-PCs-byplot)). PC2 relates to the neck and mouth constriction, from narrow to wide. PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck, from a curved to straight shape. The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods. However, the spread of shape distribution indicates a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes. In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.

```{r first-two-PCs-byplot, fig.cap= "Pottery shape distribution by each occupation phase according to the first two PCs"}

# PCA byplot
plot_PCA(op.p, ~phase, chullfilled = TRUE)
```

>>>>

```{r jpg-coordinates, eval=FALSE}
# this takes some time, so we don't run the code each time we knit, we run it once and save the output
# start shape exploration
library(Momocs)
#  it freezes on the empty ones...
jpgs_imported <- import_jpg(jpgs, threshold = 0.5)

@@

We found that 13 harmonics captured 99% of the total harmonic power in our elliptic Fourier coefficients of `r total_no` vessels from three phases. Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) shows vessel shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the most greatest differences between pre-European and Chinese periods.

```{r thin-plate-splines}
# Thin plate spline analysis (TPS)
# visualize the outline deformations required to pass from an extreme of the morphospace to the other
# Deformation grids

@@

# save figure
source(here("analysis/paper/my_tps_grid.R"))
png(here("analysis/figures/tps-iso.png"), res = 300, w = 1800, h = 1000)
par(mfrow=c(2,1)) # plot them side by side

my_tps_grid(`Pre-European`, 
         `Post-European`, 
         shp.lwd = c(2, 2), 
         shp.border = c("red", "blue"),
           amp = 1, 
         grid.size = 18, 
         legend = T)

my_tps_grid(`Pre-European`, 
         `Chinese`, 
         shp.lwd = c(2, 2), 
         shp.border = c("red", "blue"),
           amp = 1, 
         grid.size = 18, 
         legend = T)

# Isodeformation lines
dev.off()

@@

The first two principal components (PCs) of the PCA on the elliptic Fourier coefficients explain `r vari_two`% of the total variance, of which `r pc1`% is explained by the first principle component. With the third component, the first three principal components explain `r vari_three`% of the total variance. PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped (Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot)). PC2 relates to the neck and mouth constriction, from narrow to wide. PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck, from a curved to straight shape. The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods. However, the spread of shape distribution indicates a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes. In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.

```{r}
# PCA byplot, we save it to disk to use in the next chunk
png(here("analysis/figures/pca-chull.png"), res = 300, w = 1800, h = 1000)
plot_PCA(op.p, ~phase, chullfilled = TRUE)
dev.off()
```

```{r}
# composite composite image of thin plate spline and PCA output
iso_plot <- image_read(here("analysis/figures/tps-iso.png"))
iso_plot <- image_trim(iso_plot)
pca_plot <- image_read(here("analysis/figures/pca-chull.png"))

iso_plot_back <- image_scale(image_background(iso_plot, "none"), 360)
iso_pca <- image_composite(pca_plot, iso_plot_back, offset = "+50+50")
iso_pca_crop <- image_crop(iso_pca, geometry = geometry_area(1800, 750, 0, 0))
image_write(iso_pca_crop, here("analysis/figures/iso_pca_composit_crop.png"))
```

```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap="caption", fig.align='Left: Comparison of average vessel shapes across phases using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another. Right: Pottery shape distribution by each occupation phase according to the first two PCs.'}
knitr::include_graphics(here("analysis/figures/iso_pca_composit_crop.png"))
```
